Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
playing with formatting

The core of each search structure template is a \lstinline{traverse} function that navigates a data structure using the target synchronization mechanism. A template also includes wrapper functions that use \lstinline{traverse} to perform data structure operations (insert, lookup, etc.), and may include helper functions as well. The defining feature of all these functions is that they abstract away from the data structure being traversed: they operate on a generic \lstinline{node} type and take data-structure-specific functions like \lstinline{findNext} (which chooses the next node to search) as parameters. We consider lock-based synchronization in this paper, and so our generic node type can be defined in C as:
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct node_t {node *t; lock_t *lock;} node_t;
\end{lstlisting}
There are two things worth noting about this data structure. First, each node has its own lock: all of our templates will use \emph{fine-grained locking}, holding locks on as few nodes as possible to allow other parts of the data structure to be modified concurrently. Second, the \lstinline{node} struct itself is defined by the target data structure: for instance, a binary search tree might define it as 
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct node {int key; void *value; struct node_t *left, *right;} node;
\end{lstlisting}
while a linked list might use
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct node {int key; void *value; struct node_t *next;} node;
\end{lstlisting}
This is the first place where we decompose our program into a generic concurrent component and a sequential data-structure component. Next, we use \lstinline{node_t} to define template functions for specific concurrency patterns, and then prove (assuming appropriate specifications for data-structure-specific functions) that they correctly implement the data structure operations.

@@

Our implementation of \lstinline{traverse} translates the functional implementation into idiomatic C code. The lock-coupling pattern can be seen on lines 12-13, where \lstinline{traverse} acquires the next node's lock and then releases the current node's lock. The function stops when it reaches an empty leaf node (\lstinline{pn->p->t = NULL}), or when \lstinline{findNext} returns 0, indicating that the target key \lstinline{k} is in the current node. The \lstinline{traverse} function returns 1 when we reach a leaf node and 0 when we find \lstinline{k} in an existing node; the behavior of operations that call \lstinline{traverse} can vary depending on this return value. For instance, an \lstinline{insert} operation may create a new node when \lstinline{traverse} returns 1 and modify an existing node when \lstinline{traverse} returns 0, while a \lstinline{lookup} operation may fail on 1 and return the value in the current node on 0.

@@

Figure \ref{insert_lock} shows the implementation of \lstinline{insert} for the lock-coupling template.
It uses \texttt{traverse} to find the node at which to insert the key \lstinline{k}. If \texttt{traverse} returns 0, it has reached a node containing key \texttt{k}, and we only have to change the node's value to the target \texttt{value} (lines 5-6 in Figure \ref{insert_lock}). Otherwise, \texttt{traverse} has reached an empty leaf that can hold key \texttt{k}, so it calls the underlying data structure's \texttt{insertOp} function to allocate a new node with key \texttt{k} and value \texttt{v} (line 9 in Figure \ref{insert_lock}).

@@

%\wm{The fonts are still off here---for instance, p, n, and n' are mathematical variables, not references to the C code. Fix if we have time.}
The \lstinline{traverse} function can then be specified as follows:
%\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.} 
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
			\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ m \
			\right\rangle
		}\\
		\texttt{traverse(pn, k)} \\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \  n'.
			\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ 
				\\ 
				\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
				\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, \mathsf{v'}), \mathsf{range})
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
The local precondition of \lstinline{traverse} includes both the node handle $\nodeboxrep(n)$ and its contents $\mathsf{R}(n)$, indicating that a thread should already hold $n$'s lock before calling \lstinline{traverse}. The output of \lstinline{traverse} is a new node $n'$ in the data structure such that the key \texttt{k} falls within the range of $n'$. The local postcondition then includes the handle of the new node $\nodeboxrep(n')$, its contents $\mathsf{R}(n')$, and a Boolean variable $\mathit{res}$ indicating whether \lstinline{traverse} found an empty node or a node with key \lstinline{k}.

The resources $\mathsf{R}$ contained in a node depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. Formally, the lock invariant is defined by 
\begin{align*}
	&\mathsf{node\_contents}(n, \mathsf{c}, \mathsf{range}) \triangleq \mathsf{ghost\_node(n, \mathsf{c}, range)} \ \ast \ \mathsf{node\_data}(n, \mathsf{c})\\
	&\mathsf{R}(n) \triangleq \exists\ \mathsf{c}, \mathsf{range}.\ \mathsf{node\_contents}(n, \mathsf{c}, \mathsf{range})\end{align*}
where the definition of $\mathsf{node\_data}$ is supplied by the target data structure. The contents $\mathsf{c}$ of a node can be either a key-value pair $(\texttt{k}, \mathsf{v})$, or the empty contents $\cdot$ (used for nodes that have been allocated but not yet assigned keys).

@@

The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ p, n, \mathsf{c}, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \  \nodeboxrep(n)  \ \ast \\  &\mathsf{node\_contents}(n, \mathsf{c}, \mathsf{range})
\end{align*}

\begin{figure}[!ht]
	$\color{blue}
	\forall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
	\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ m \
	\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  n'.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ 
		\\ 
		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, \mathsf{v'}), \mathsf{range})
	\end{array}
	\ \Bigg| \ \treerep\ m \
	\right\rangle$
	\caption{Proof outline of the lock-coupling \texttt{traverse} function}
	\label{proof_lock_traverse}
\end{figure}

Figure \ref{proof_lock_traverse} shows the proof outline of the \texttt{traverse} function.
We begin by checking whether the current node is null (line 5 of Figure~\ref{proof_lock_traverse}); if it is, we have found the leaf where \lstinline{k} belongs, and can prove the postcondition with $\mathit{res} = \texttt{true}$. Otherwise, we pass the $\mathsf{node\_data}$ from the lock invariant to \lstinline{findNext} (line 11), which returns a new node $n''$ to visit, stored in \lstinline{pn->n}. If \lstinline{findNext} returns 0, we have found a node containing \lstinline{k}, and can prove the postcondition with $\mathit{res} = \mathtt{false}$. Otherwise, we acquire $n\texttt{->lock}$, gaining access to the resources $\mathsf{R}(n)$, and then release the lock of the current node $p$ and return its resources. %It is important to note that \texttt{p} point to the same memory address as \texttt{n} (line 4 in Figure \ref{traverse_lock_b}).
By acquiring the lock for the next node $n$ before releasing the lock for the current node $p$, we ensure that the connection between the two nodes remains intact and unaltered while we traverse it, and re-establish $\mathsf{traverse\_inv}$ for the new values in \lstinline{pn}. 
%The proof concludes by verifying that the \lstinline{traverse} function fulfills the post-condition of the specification mentioned earlier.
The two \lstinline{return}s are also the two possible linearization points of the function.

@@

The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle
			\begin{array}{c}
				\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ m
			\end{array}
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \  n'.
			\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ 
				\\ 
				\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
				\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, \mathsf{v'}), \mathsf{range})
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
As before, the \texttt{traverse} function will navigate the structure and return a node $n'$ whose range includes \texttt{k}, as well as acquiring $n'$'s lock and returning its contents. Also as before, the Boolean $\mathit{res}$ indicates whether $n'$ is empty or contains the key \lstinline{k}.

@@

With this invariant in hand, the proof of \texttt{traverse} proceeds as follows. We begin by using the $\infp$ predicate to acquire the lock on \lstinline{pn->n} (line 4 in Figure \ref{traverse_giveup_b}). We then call \lstinline{inRange} to check whether we are still on the path to a node that can hold \lstinline{k}. If the check fails (lines 18-20), we immediately release the lock and start over from the root node $r$. Otherwise, we proceed as in the lock-coupling template: if the current node's contents are \texttt{NULL}, we have found the empty leaf where \texttt{k} belongs (and can satisfy the postcondition); if \texttt{findNext} returns 0, we have found the node containing \texttt{k} (and can satisfy the postcondition); otherwise, we release the lock and re-establish $\mathsf{traverse\_inv}$ for the new node indicated by \texttt{findNext}. In this case, we once again hold no locks, and have only the $\infp$ assertion indicating that the new node is in the data structure. % and add the additional predicate $\texttt{k} \in \texttt{range(n)}$ in the \texttt{then} of \texttt{inRange}. Similar to the proof of the lock-coupling template's \texttt{traverse}, the traversal stops when it reaches the base case, which occurs when the current node is \texttt{NULL} (line 8 in Figure \ref{traverse_giveup_b}). At this point, we can prove the post-condition using the specification, which indicates that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$.

@@

\begin{figure}[!ht]
	$\color{blue}
	\forall \  m. \left\langle
	\begin{array}{c}
		\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \nodeboxrep(\texttt{n}) \ \big| \ \treerep\ m
	\end{array}
	\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
	\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \nodeboxrep(\texttt{n'}) \ \ast \ 
		\\ 
		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(\texttt{n'}, \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(\texttt{n'}, (\mathsf{k}, \mathsf{v'}), \mathsf{range})
	\end{array}
	\ \Bigg| \ \treerep\ m \
	\right\rangle$
	\caption{Proof outline of the give-up \texttt{traverse} function}
	\label{proof_giveup_traverse}
\end{figure}

>>>>

The core of a search structure template is a \lstinline{traverse} function that navigates a data structure using the chosen synchronization mechanism. A template also includes wrapper functions that use \lstinline{traverse} to perform data structure operations (insert, lookup, etc.), and may include helper functions as well. The defining feature of all these functions is that they abstract away from the data structure being traversed: they operate on a generic \lstinline{node} type and take data-structure-specific functions like \lstinline{findNext} (which chooses the next node to search) as parameters. We consider lock-based synchronization in this paper, and so our generic node type can be defined in C as:
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct node_t {node *t; lock_t *lock;} node_t;
\end{lstlisting}
There are two things worth noting about this data structure. First, each node has its own lock: all of our templates will use \emph{fine-grained locking}, holding locks on as few nodes as possible to allow other parts of the data structure to be modified concurrently. Second, the \lstinline{node} struct itself is defined by the target data structure: for instance, a binary search tree might define it as 
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct node {int key; void *value; struct node_t *left, *right;} node;
\end{lstlisting}
while a linked list might use
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct node {int key; void *value; struct node_t *next;} node;
\end{lstlisting}
This is the first place where we decompose our program into a generic concurrent component and a sequential data-structure component. Next, we use \lstinline{node_t} to define template functions for specific concurrency patterns, and then prove (assuming appropriate specifications for data-structure-specific functions) that they correctly implement the data structure operations.

@@

Our implementation of \lstinline{traverse} translates the functional implementation into idiomatic C code. The lock-coupling pattern can be seen on lines 12-13, where \lstinline{traverse} acquires the next node's lock and then releases the current node's lock. The function stops when it reaches an empty node (\lstinline{pn->p->t == NULL}), or when \lstinline{findNext} returns 0, indicating that the target key \lstinline{k} is in the current node. The \lstinline{traverse} function returns 1 when we reach an empty node and 0 when we find \lstinline{k} in an existing node; the behavior of operations that call \lstinline{traverse} can vary depending on this return value. For instance, an \lstinline{insert} operation may create a new node when \lstinline{traverse} returns 1 and modify an existing node when \lstinline{traverse} returns 0, while a \lstinline{lookup} operation may fail on 1 and return the value in the current node on 0.

@@

Figure \ref{insert_lock} shows the implementation of \lstinline{insert} for the lock-coupling template.
It uses \texttt{traverse} to find the node at which to insert the key \lstinline{k}. If \texttt{traverse} returns 0, it has reached a node containing key \texttt{k}, and we only have to change the node's value to the target \texttt{value} (lines 5-6 in Figure \ref{insert_lock}). Otherwise, \texttt{traverse} has reached an empty node that can hold key \texttt{k}, so it calls the underlying data structure's \texttt{insertOp} function to allocate a new node with key \texttt{k} and value \texttt{v} (line 9 in Figure \ref{insert_lock}).

@@

%\wm{The fonts are still off here---for instance, p, n, and n' are mathematical variables, not references to the C code. Fix if we have time.}
The \lstinline{traverse} function can then be specified as follows:
%\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.} 
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
			\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ m \
			\right\rangle
		}\\
		\texttt{traverse(pn, k)} \\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \  n'.
			\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
				\\ 
				\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
				\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
The local precondition of \lstinline{traverse} includes both the node handle $\nodeboxrep(n)$ and its contents $\mathsf{R}(n)$, indicating that a thread should already hold $n$'s lock before calling \lstinline{traverse}. The output of \lstinline{traverse} is a new node $n'$ in the data structure such that the key \texttt{k} falls within the range of $n'$. The local postcondition then includes the handle of the new node $\nodeboxrep(n')$, its contents $\mathsf{R}(n')$, and a Boolean variable $\mathit{res}$ indicating whether \lstinline{traverse} found an empty node or a node with key \lstinline{k}.

The resources $\mathsf{R}$ contained in a node depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. Formally, the lock invariant is defined by 
\begin{align*}
	&\mathsf{node\_contents}(n, c, \mathsf{range}) \triangleq \mathsf{ghost\_node}(n, c,\mathsf{range}) \ \ast \ \mathsf{node\_data}(n, c)\\
	&\mathsf{R}(n) \triangleq \exists\ c, \mathsf{range}.\ \mathsf{node\_contents}(n, c, \mathsf{range})\end{align*}
where the definition of $\mathsf{node\_data}$ is supplied by the target data structure. The contents $c$ of a node can be either a key-value pair $(k, v)$, or the empty contents $\cdot$ (used for nodes that have been allocated but not yet assigned keys).

@@

The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ p, n, c, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \  \nodeboxrep(n)  \ \ast \\  &\mathsf{node\_contents}(n, c, \mathsf{range})
\end{align*}

\begin{figure}[!ht]
	$\color{blue}
	\forall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
	\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ m \
	\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  n'.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast\ 
		\\ 
		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
	\end{array}
	\ \Bigg| \ \treerep\ m \
	\right\rangle$
	\caption{Proof outline of the lock-coupling \texttt{traverse} function}
	\label{proof_lock_traverse}
\end{figure}

Figure \ref{proof_lock_traverse} shows the proof outline of the \texttt{traverse} function.
We begin by checking whether the current node is null (line 5 of Figure~\ref{proof_lock_traverse}); if it is, we have found the empty node where \lstinline{k} belongs, and can prove the postcondition with $\mathit{res} = \mathsf{true}$. Otherwise, we pass the $\mathsf{node\_data}$ from the lock invariant to \lstinline{findNext} (line 11), which returns a new node $n''$ to visit, stored in \lstinline{pn->n}. If \lstinline{findNext} returns 0, we have found a node containing \lstinline{k}, and can prove the postcondition with $\mathit{res} = \mathsf{false}$. Otherwise, we acquire $n\texttt{->lock}$, gaining access to the resources $\mathsf{R}(n)$, and then release the lock of the current node $p$ and return its resources. %It is important to note that \texttt{p} point to the same memory address as \texttt{n} (line 4 in Figure \ref{traverse_lock_b}).
By acquiring the lock for the next node $n$ before releasing the lock for the current node $p$, we ensure that the connection between the two nodes remains intact and unaltered while we traverse it, and re-establish $\mathsf{traverse\_inv}$ for the new values in \lstinline{pn}. 
%The proof concludes by verifying that the \lstinline{traverse} function fulfills the post-condition of the specification mentioned earlier.
The two \lstinline{return}s are also the two possible linearization points of the function.

@@

The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle
			\begin{array}{c}
				\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ m
			\end{array}
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \  n'.
			\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ \texttt{k} \in \mathsf{range}\ \ast\ 
				\\ 
				\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
				\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
As before, the \texttt{traverse} function will navigate the structure and return a node $n'$ whose range includes \texttt{k}, as well as acquiring $n'$'s lock and returning its contents. Also as before, the Boolean $\mathit{res}$ indicates whether $n'$ is empty or contains the key \lstinline{k}.

@@

With this invariant in hand, the proof of \texttt{traverse} proceeds as follows. We begin by using the $\infp$ predicate to acquire the lock on \lstinline{pn->n} (line 4 in Figure \ref{traverse_giveup_b}). We then call \lstinline{inRange} to check whether we are still on the path to a node that can hold \lstinline{k}. If the check fails (lines 18-20), we immediately release the lock and start over from the root node $r$. Otherwise, we proceed as in the lock-coupling template: if the current node's contents are \texttt{NULL}, we have found the empty node where \texttt{k} belongs (and can satisfy the postcondition); if \texttt{findNext} returns 0, we have found the node containing \texttt{k} (and can satisfy the postcondition); otherwise, we release the lock and re-establish $\mathsf{traverse\_inv}$ for the new node indicated by \texttt{findNext}. In this case, we once again hold no locks, and have only the $\infp$ assertion indicating that the new node is in the data structure. % and add the additional predicate $\texttt{k} \in \texttt{range(n)}$ in the \texttt{then} of \texttt{inRange}. Similar to the proof of the lock-coupling template's \texttt{traverse}, the traversal stops when it reaches the base case, which occurs when the current node is \texttt{NULL} (line 8 in Figure \ref{traverse_giveup_b}). At this point, we can prove the post-condition using the specification, which indicates that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$.

@@

\begin{figure}[!ht]
	$\color{blue}
	\forall \  m. \left\langle
	\begin{array}{c}
		\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \nodeboxrep(\texttt{n}) \ \big| \ \treerep\ m
	\end{array}
	\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
	\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \nodeboxrep(\texttt{n'}) \ \ast \ \texttt{k} \in \mathsf{range}\ \ast \ 
		\\ 
		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(\texttt{n'}, \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(\texttt{n'}, (\mathsf{k}, v), \mathsf{range})
	\end{array}
	\ \Bigg| \ \treerep\ m \
	\right\rangle$
	\caption{Proof outline of the give-up \texttt{traverse} function}
	\label{proof_giveup_traverse}
\end{figure}
